% ALGORITHMS FOR MODULAR ELLIPTIC CURVES       Last change: 14/8/96
%
% BIBLIOGRAPHY
%
% Use AmSTeX 2.0
%
\input book.def
\advance\pageno by\chaponepages
\advance\pageno by\chaptwopages
\advance\pageno by\appendixpages
\advance\pageno by\chapthreepages
\advance\pageno by\chapfourpages
\advance\pageno by\tabonepages
\advance\pageno by\tabtwopages
\advance\pageno by\tabthreepages
\advance\pageno by\tabfourpages
\advance\pageno by\tabfivepages
%
\advance\vsize by 5pt
\topmatter
%
\title Bibliography \endtitle
%
\endtopmatter
%
%
\Refs\nofrills{{}}
\widestnumber\no{999}

\ref \no \AtkinL \by A.~O.~L.~Atkin and J.~Lehner
\paper Hecke operators on $\Gamma_0(m)$
\jour Math.~Ann. \vol 185 \yr 1970 \pages 134--160
\endref\smallskip

\ref \no \Antwerp \by B.~J.~Birch and W.~Kuyk (eds.)
\book Modular Functions of One Variable IV
\bookinfo Lecture Notes in  Mathematics 476
\publ Springer-Verlag  \yr 1975
\endref\smallskip

\ref \no \BirchSD \by B.~J.~Birch and H.~P.~F.~Swinnerton-Dyer
\paper Notes on elliptic curves.~I.
\jour J.~Reine Angew. Math. \vol 212 \yr 1963 \pages 7--25
\endref\smallskip

\ref \no \BrumerK \by A.~Brumer and K.~Kramer
\paper The rank of elliptic curves
\jour Duke Math.~J. \vol 44 \yr 1977 \pages 715--743
\endref\smallskip

\ref \no \BuhlerG \by J.~P.~Buhler and B.~H.~Gross
\paper Arithmetic on curves with complex multiplication.~II
\jour Invent.~Math. \vol 79 \yr 1985 \pages  11--29
\endref\smallskip

\ref \no \BuhlerGZ \by J.~P.~Buhler, B.~H.~Gross and D.~B.~Zagier
\paper On the conjecture of Birch and Swinnerton Dyer for an elliptic
curve of rank $3$
\jour  Math.~Comp. \vol 44 \yr 1985 \pages 473--481
\endref\smallskip

\ref \no \Carayol \by H.~Carayol
\paper Sur les repr\'esentations $l$-adiques attach\'ees aux formes modulaires
de Hilbert
\jour C.~R.~Acad.~Sc.~Paris \vol 296 \yr 1983 \page 629
\endref\smallskip

\ref \no \Cassels \by J.~W.~S.~Cassels
\book Elliptic curves
\bookinfo LMS Student Texts
\publ Cambridge University Press \yr 1991
\endref\smallskip

\ref \no \Cohenbook \by H.~Cohen
\book A Course in Computational Algebraic Number Theory
\bookinfo Graduate Texts in Mathematics 138
\publ Springer-Verlag \yr 1993
\endref\smallskip

\ref \no \Connell \by I.~Connell
\book Notes on elliptic curves
\bookinfo (Unpublished lecture notes, McGill University)
\yr 1991
\endref\smallskip

\ref \no \Cox \by D.~Cox
\paper The arithmetic-geometric mean of Gauss 
\jour l'Enseignement Math. \vol 30 \yr 1984 \pages 270--330
\endref\smallskip

\ref \no \JECa \by J.~E.~Cremona
\paper Hyperbolic tessellations, modular symbols, and elliptic curves over
complex quadratic fields
\jour Compositio Math. \vol 51 \yr 1984 \pages 275--323
\endref\smallskip

\ref \no \JECb \by J.~E.~Cremona \paper Hyperbolic tessellations, modular
symbols, and elliptic curves over complex quadratic fields (Addendum and
Errata) \jour Compositio Math. \vol 63 \yr 1987 \pages 271--272
\endref\smallskip

\ref \no \gammaone \by J.~E.~Cremona \paper Modular symbols for 
$\Gamma_1(N)$ and elliptic curves with everywhere goood reduction
\jour Math.~Proc.~Camb.~Phil.~Soc. \vol 111 \yr 1992 \pages 199--218
\endref\smallskip

\ref \no \bsdiqf \by J.~E.~Cremona and E.~Whitley \paper Periods of cusp
forms and elliptic curves over imaginary quadratic fields 
\jour Math.~Comp. \vol 62 \yr 1994 \pages 407--429
\endref\smallskip

\ref \no \JCsha \by J.~E.~Cremona 
\paper The analytic order of $\Sha$ for modular elliptic curves
\jour J.~de Th\'eorie des Nombres de Bordeaux \vol 5 \yr 1993 \pages 179--184
\endref\smallskip

\ref \no \JCdegphi \by J.~E.~Cremona 
\paper Computing the degree of the modular parametrization of a
modular elliptic curve
\jour Math.~Comp. \vol 64 \yr 1995 \pages 1235--1250
\endref\smallskip

\ref \no \JCperiods \by J.~E.~Cremona 
\paper Computing periods of cusp forms and modular elliptic curves
\jour Exp.~Math. \vol 6:2 \yr 1997 \pages 97--107
\endref\smallskip

\ref \no \JCPS \by J.~E.~Cremona and P.~Serf
\paper Computing the rank of elliptic curves over real quadratic
number fields of class number $1$
\jour University of Exeter Department of Mathematics Preprint
\vol M96/14 \yr 1996
\endref\smallskip

\ref \no \JCinvariants \by J.~E.~Cremona 
\paper Classical invariants and $2$-descent on elliptic curves
\jour University of Exeter Department of Mathematics Preprint
\vol M96/16 \yr 1996
\endref\smallskip

\ref \no \Edixhoven \by B.~Edixhoven
\paper On the Manin constants of modular elliptic curves
\pages 25--39
\inbook Arithmetic Algebraic Geometry
\bookinfo Progress in Mathematics {\bf89}
\publ Birkhauser \yr 1991
\endref\smallskip

\ref \no \Faltings \by G.~Faltings
\paper Endlichkeitss\"atze f\"ur abelsche Variet\"aten \"uber Zahlk\"orpern
\jour Invent.~Math. \vol 73 \yr 1983 \pages 349--366
\endref\smallskip

\ref \no \Ferm \by S.~Fermigier
\paper Construction of high rank elliptic curves over $\Q(T)$ and $\Q$
with nontrivial $2$-torsion
\jour C.~R.~Acad.~Sc.~Paris \vol 322 \yr 1996 \pages 949--952
\endref\smallskip

\ref \no \Luiz \by L.~Figueiredo
\book Serre's conjecture over imaginary quadratic fields
\bookinfo PhD~thesis, University of Cambridge \yr 1995
\endref\smallskip

\ref \no \GrossZagiera \by B.~H.~Gross and D.~Zagier
\paper Points de Heegner et d\'eriv\'ees de fonctions $L$
\jour C.~R.~Acad.~Sc. Paris \vol 297 \yr 1983 \pages 85--87
\endref\smallskip

\ref \no \GrossZagierb \by B.~H.~Gross and D.~Zagier
\paper Heegner points and derivatives of $L$-series
\jour Invent.~Math. \vol 84 \yr 1986 \pages 225--320
\endref\smallskip

%% not referred to in 2nd edn.
%
%\ref \no \HaradaLang \by K.~Harada and M.~L.~Lang
%\paper Some elliptic curves arising from the Leech lattice
%\jour Journal of Algebra \vol 125 \yr 1989 \pages 298--310
%\endref\smallskip


\ref \no \Husemoller \by D.~Husem\"oller
\book Elliptic curves
\publ Springer-Verlag  \yr 1987
\endref\smallskip

\ref \no \Knapp \by A.~W.~Knapp
\book Elliptic Curves
\bookinfo Mathematical Notes {\bf40}
\publ Princeton University Press
\yr 1992
\endref\smallskip

\ref \no \Kolyvagin \by V.~I.~Kolyvagin
\paper Finiteness of $E(\Q)$ and $\Sha_{E/\Q}$ for a subclass of Weil
curves
\jour Math.~USSR Izvest. \vol 32 \yr 1989 \pages 523--542
\endref\smallskip

\ref \no \Kraus \by A.~Kraus
\paper Quelques remarques \`a propos des invariants $c_4$, $c_6$ et 
$\Delta$ d'une courbe elliptique
\jour Acta Arith.~\yr 1989 \vol 54 \pages 75--80
\endref\smallskip

\ref \no \Langa \by S.~Lang
\book Elliptic functions
\publ Addison-Wesley \yr 1973
\endref\smallskip

\ref \no \Langb \by S.~Lang
\book Introduction to modular forms
\publ Springer-Verlag \yr 1976
\endref\smallskip

\ref \no \Langc \by S.~Lang
\book Elliptic curves: diophantine analysis
\publ Springer-Verlag  \yr 1976
\endref\smallskip

\ref \no \Laska \by M.~Laska
\paper An algorithm for finding a minimal Weierstrass equation for an 
elliptic curve
\jour Math.~Comp. \yr 1982 \vol 38 \pages 257--260
\endref\smallskip

\ref \no \Laskabook \by M.~Laska
\book Elliptic curves over number fields with prescribed reduction type
\publ Vieweg \bookinfo Aspects of Mathematics Vol.~E4 \yr 1983 
\endref\smallskip

%% not referred to in 2nd edn.
%
%\ref \no \Livne \by R.~Livne
%\paper Cubic exponential sums and Galois representations
%\jour Contemporary Mathematics  \yr 1987 \vol 67 \pages 247--261
%\endref\smallskip

\ref \no \Lutz \by E.~Lutz
\paper Sur l'\'equation $y^2=x^3-Ax-B$ dans les corps $p$-adic
\jour J.~Reine Angew.~Math. \vol 177 \yr 1937 \pages 237--247
\endref\smallskip

\ref \no \Manin \by Ju.~I.~Manin
\paper Parabolic points and zeta-functions of modular curves
\jour Math.~USSR-Izv. \vol 6 \yr 1972 \pages 19--64
\endref\smallskip

\ref \no \Mazur \by B.~Mazur
\paper Courbes elliptiques et symboles modulaires
\jour S\'eminaire Bourbaki No.~414 \yr 1972 \pages 277--294
\endref\smallskip

\ref \no \MazurTa \by B.~Mazur
\paper Modular curves and the Eisenstein ideal
\jour IHES Publ.~Math. \vol 47 \yr 1977 \pages 33--186
\endref\smallskip

\ref \no \MazurTb \by B.~Mazur
\paper Rational isogenies of prime degree
\jour Invent.~Math. \vol 44 \yr 1978 \pages 129--162
\endref\smallskip

\ref \no \MazurSD \by B.~Mazur and H.~P.~F.~Swinnerton-Dyer
\paper Arithmetic of Weil curves
\jour Invent.~Math. \vol 25 \yr 1974 \pages 1--61
\endref\smallskip

\ref \no \Merel \by L.~Merel
\paper Op\'erateurs de Hecke pour $\Gamma_0(N)$ et fractions continues
\jour Annales de l'Institut Fourier de l'Uni\-ver\-sit\'e de Grenoble \vol
41 no.~3 \yr 1991 \pages 519--537
\endref\smallskip

\ref \no \Mestre \by J.-F.~Mestre
\paper La m\'ethode des graphes. Exemples et applications.
\inbook Proceedings of the International Conference on Class Numbers and
Units of Algebraic Number Fields (Y.~Yama\-moto and H.~Yokoi, eds.)
\publ Katata \yr 1986
\pages 217--242
\endref\smallskip

\ref \no \Murty \by M.~R.~Murty and V.~K.~Murty
\paper Mean values of derivatives of modular $L$-series
\jour Ann.~Math. \vol 133 \yr 1991 \pages 447--475
\endref\smallskip

\ref \no \Nagao \by K.~Nagao
\paper An example of an elliptic curve over $\Q$ with rank $\ge20$
\jour Proceedings of the Japan Academy Series A
\vol 69 \yr 1993 \pages 291--293
\endref\smallskip

\ref \no \Nag \by T.~Nagell
\paper Solution de quelque probl\`emes dans la th\'eorie
arithm\'etique des cubiques planes du premier genre
\jour Wid.~Akad.~Skrifter Oslo I \yr 1935
\endref\smallskip

\ref \no \Neron \by A.~N\'eron
\paper Mod\`eles minimaux des vari\'et\'es ab\'eliennes sur les corps
locaux et globeaux
\jour IHES Publ.~Math. \vol 21 \yr 1964 \pages 361--482
\endref\smallskip

\ref \no \Pincha \by R.~G.~E.~Pinch
\book Elliptic curves over number fields
\bookinfo DPhil~thesis, University of Oxford \yr 1982
\endref\smallskip

\ref \no \Pinchb \by R.~G.~E.~Pinch
\paper Elliptic curves with everywhere good reduction
\jour (preprint)
\endref\smallskip

\ref \no \PZass \by  M.~Pohst and H.~Zassenhaus
\book Algorithmic Algebraic Number Theory
\bookinfo Encyclopedia of mathematics and its applications
\publ Cambridge University Press \yr 1989
\endref\smallskip

%% not referred to in 2nd edn.
%
%\ref \no \Scholl \by A.~J.~Scholl
%\paper The $l$-adic representations attached to a certain non\-congruence
%subgroup
%\jour J.~Reine Angew.~Math. \vol 392 \yr 1988 \pages 1--15
%\endref\smallskip

\ref \no \Schoof \by R.~Schoof
\paper Elliptic curves over finite fields and the computation of
square roots mod $p$
\jour Math.~Comp. \vol 44 \yr 1985 \pages 483--494
\endref\smallskip

\ref \no \PSthesis \by P.~Serf
\book The rank of elliptic curves over real quadratic number fields of
class number $1$
\bookinfo PhD thesis, Universit\"at des Saarlandes, Saarbr\"ucken
\yr 1995
\endref\smallskip

\ref \no \Serre \by J.-P.~Serre
\paper Propri\'et\'es galoisiennes des points d'ordre fini des courbes
elliptiques
\jour Invent.~Math. \vol 15 \yr 1972 \pages 259--331
\endref\smallskip

\ref \no \SerreNotes \by J.-P.~Serre
\book R\'esum\'e de cours \bookinfo Annuaire du Coll\`ege de France
\yr 1984--85 
\endref\smallskip

\ref \no \Shimura \by  G.~Shimura
\book Introduction to the arithmetic theory of automorphic functions
\publ Math.~Soc. Japan No.~11 \yr 1971
\endref\smallskip

\ref \no \SSthesis \by S.~Siksek
\book Descent on curves of genus one
\bookinfo PhD thesis, University of Exeter
\yr 1995
\endref\smallskip

\ref \no \SSpaper \by S.~Siksek
\paper Infinite Descent on Elliptic Curves
\jour Rocky Mountain Journal of Mathematics
\vol 25 no.~4 \yr 1995 \pages 1501--1538
\endref\smallskip

\ref \no \Silvera \by J.~H.~Silverman
\book The Arithmetic of Elliptic Curves
\bookinfo GTM 106
\publ Springer-Verlag  \yr 1986
\endref\smallskip

\ref \no \Silverb \by J.~H.~Silverman
\paper Computing heights on elliptic curves
\jour Math.~Comp. \vol 51 \yr 1988 \pages 339--358
\endref\smallskip

\ref \no \Silverc \by J.~H.~Silverman
\paper The difference between the Weil height and the canonical height on 
elliptic curves
\jour Math.~Comp. \vol 55 \yr 1990 \pages 723--743
\endref\smallskip

\ref \no \Silverd \by J.~H.~Silverman
\book Advanced Topics in the Arithmetic of Elliptic Curves
\bookinfo GTM 151
\publ \break Springer-Verlag  \yr 1994
\endref\smallskip

\ref \no \Silvere \by J.~H.~Silverman
\paper Computing canonical heights with little (or no) factorization
\jour Math. Comp. \vol 66 \yr 1997 \pages 787--805
\endref\smallskip

\ref \no \Nigeletal \by N.~P.~Smart, S.~Siksek and J.~R.~Merriman
\paper Explicit $4$-descents on an elliptic curve
\jour Acta Arith. \yr 1996 \vol LXXVII.4 \pages 385--404
\endref\smallskip

\ref \no \SDBirch \by H.~P.~F.~Swinnerton Dyer and B.~J.~Birch
\paper Elliptic curves and modular functions
\inbook Modular Functions of One Variable IV
\bookinfo Lecture Notes in  Mathematics 476
\publ Springer-Verlag \yr 1975
\endref\smallskip

\ref \no \Tate \by J.~Tate
\paper Algorithm for determining the singular fiber in an elliptic pencil
\inbook Modular Functions of One Variable IV
\bookinfo Lecture Notes in  Mathematics 476
\publ Springer-Verlag \yr 1975
\endref\smallskip

\ref \no \Tateht \by J.~Tate
\finalinfo Letter to J.-P.~Serre, Oct.~1, 1979
\endref\smallskip

\ref \no \Tingley \by D.~J.~Tingley
\book Elliptic curves uniformized by modular functions
\bookinfo DPhil~thesis, University of Oxford \yr 1975
\endref\smallskip

\ref \no \Velu \by J.~V\'elu
\paper Isog\'enies entre courbes elliptiques
\jour C.~R.~Acad.~Sc.~Paris \vol 273 \yr 1971 \pages 238--241
\endref\smallskip

\ref \no \Zagier \by D.~Zagier
\paper Modular Parametrizations of Elliptic Curves
\jour Canadian Math.~Bull. \yr 1985  \vol 28 \pages 372--384
\endref\smallskip

\ref \no \Zimmer \by J.~Gebel and H.~G.~Zimmer 
\paper Computing the Mordell-Weil Group of an Elliptic Curve over $\Q$
\inbook Elliptic Curves and Related Topics 
(H.~Kisilevsky and M.~Ram Murty, eds.)  
\bookinfo CRM Proceedings and Lecture Notes 
\yr 1994 \vol 4 \pages 61--83 
\endref\smallskip


%
%
\endRefs
%
%\enddocument
\end
